---
title: "Introduction to Data pre-processing Part4: Paracetamol Study"
author: "Reika Masuda"
date: "13th Feb 2025"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    number_sections: false
    theme: lumen
editor_options: 
  chunk_output_type: inline
---

# load the required packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,results = 'asis')

# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("ropls")

# 
# devtools::load_all("C:Download/nmr-peaks/nmr-peaks")
# devtools::load_all("C:Download/nmr-spectra-processing/nmr-spectra-processing")
# devtools::load_all("C:Download/nmr-parser/nmr-parser")
# devtools::load_all("C:Download/mva-plots/mva-plots")

library(tidyverse)
library(dplyr)
library(ggplot2)
```


## Dummy Metadata corresponding to the Urine NMR data

```{r}
MetaData<-data.frame(SourceID = c(1,2,3,4,5),
                     Paracetamol = c(1,1,1,1,1),
                     Sex = c("M","F","F","M","M"))
```

## load the Urine 1D spectra data

```{r}
# Set the folder location to where the Bruker NMR data is
folder <- "D:/NMR Course 2025/NMR_short_course_Paracetamol/NMR_short_course_Paracetamol/"

## load in the RunOrder
RunOrder<-read.csv("D:/NMR Course 2025/NMR_short_course_Paracetamol/NMR_short_course_Paracetamol_RunOrder.csv")

# find the expno that are corresponding to 1D spectra data
expno<-RunOrder$EXPNO[which(RunOrder$EXPERIMENT=="PROF_URINE_NOESY")]

# read the multiple 1D spectra data (with eretic factor correction)
for (i in 1:length(expno)) {
  # cat("\n", "reading the spectra data of: ", expno[i], "\n")
  test_spec <- readSpectrum(paste0(folder, expno[i]))
  test_eretic <- readEretic(paste0(folder, expno[i], "/QuantFactorSample.xml"))
  
  # cat("\n", "eretic factor of: ", test_eretic$ereticFactor, "\n")
  if (i == 1) {
    spec <- test_spec$spec$y / test_eretic$ereticFactor
    ppm <- test_spec$spec$x
    ID<-readTitle(paste0(folder, expno[i],"/pdata/1/title"))$value
  } else{
    spec <- rbind(spec, test_spec$spec$y / test_eretic$ereticFactor)
    ID<-c(ID,readTitle(paste0(folder, expno[i],"/pdata/1/title"))$value)
  }
  rm(test_spec, test_eretic)
}


## clean up the ID names
spec_anno<-data.frame(title=ID)%>%
  mutate(title = gsub("TITLE ","",title),
         sourceID = sapply(strsplit(title, "_"), "[", 2),
         timepoint = sapply(strsplit(title, "_"), "[", 3))

colnames(spec) <- ppm
rownames(spec) <- spec_anno$title

matspec(spec, ppm, roi = c(1.46,1.50),interactive = T)

```

## load the IVDr small molecule reports

```{r}

for (i in 1:length(expno)) {
  # Read in the quant report xml
  quant_data <- readQuant(paste0(folder, expno[i], "/pdata/1/urine_quant_report_e.xml"))$data
  
  if (i == 1) {
    # Create a data frame to store the rowConc (row concentration values) and sigCorr (signal correction values)
    quant <- data.frame(variable = quant_data$name, as.numeric(quant_data$rawConc))
    sig <- data.frame(variable = quant_data$name, as.numeric(quant_data$sigCorr))
    ID<-readTitle(paste0(folder, expno[i],"/pdata/1/title"))$value
  } else {
    # Append the quant, sig data to the data frame
    quant <- cbind(quant, as.numeric(quant_data$rawConc))
    sig <- cbind(sig, as.numeric(quant_data$sigCorr))
    ID<-c(ID,readTitle(paste0(folder, expno[i],"/pdata/1/title"))$value)
  }
  
  rm(quant_data)
}


# Transpose the data frame
quant <- data.frame(t(quant))
colnames(quant) <- quant[1, ]
quant <- quant[-1, ]

sig <- data.frame(t(sig))
colnames(sig) <- sig[1, ]
sig <- sig[-1, ]


## clean up the ID names
quant_anno<-data.frame(title=ID)%>%
  mutate(title = gsub("TITLE ","",title),
         sourceID = sapply(strsplit(title, "_"), "[", 2),
         timepoint = sapply(strsplit(title, "_"), "[", 3))



# Add the experiment number to the row names
rownames(sig) <- quant_anno$title
rownames(quant) <- quant_anno$title

# Convert the data frame variables to numeric
sig<-data.frame(apply(sig,2,as.numeric))
quant<-data.frame(apply(quant,2,as.numeric))

## Define the metabolites of interest
metabolite = c("Paracetamol","Paracetamol.glucuronide","Caffeine","Trigonelline","Creatine","Creatinine","Alanine","D.Glucose"  )

cat("\n", "we have selected following metabolites of interest: ",paste0(metabolite,collapse = ", "),"\n")


## Subset the data to the metabolites of interest
quant<-quant[,which(colnames(quant)%in%metabolite)]
sig<-sig[,which(colnames(sig)%in%metabolite)]

htmlTable::htmlTable(sig,caption = "The lineshpe fit correlation values of the metabolites of interest")
```

## Matching the NMR and IVDr data based on the Title, match with MetaData

```{r}
# Match the NMR and IVDr data based on the Title
idx<-match(spec_anno$title,quant_anno$title)
quant<-quant[idx,]
quant_anno<-quant_anno[idx,]
table(spec_anno$title == quant_anno$title)
rm(idx)

# we need to match the metadata with the NMR and IVDr data
idx<-match(quant_anno$sourceID,MetaData$SourceID)
MetaData<-MetaData[idx,]
MetaData$timepoint<-quant_anno$timepoint
cat("\n","The row number of MetaData increased from 5 to ", nrow(MetaData),"\n")
rm(idx)

```

## PCA analysis of IVDr Data (targeted approach)

```{r}
mod_ivdr<-PCA(quant)

plotScores(model = mod_ivdr,optns = list(color = MetaData$Sex,shape = MetaData$timepoint))
plotScores(model = mod_ivdr,optns = list(color = factor(MetaData$SourceID),shape = MetaData$timepoint,colorTitle = "sourceID",shapeTitle = "timepoint"))

```

## track the longitudinal changes of the metabolites

```{r}
data.frame(MetaData%>%
             mutate(timepoint = factor(timepoint,
                                       levels = c("baseline","2hours","4hours","6hours"))),quant)%>%
  pivot_longer(cols = Creatinine:D.Glucose,names_to = "metabolite",values_to = "val")%>%
  ggplot(aes(x = timepoint, y = val, group = factor(SourceID), colour = factor(SourceID)))+
  geom_point()+
  geom_line()+
  facet_wrap(~metabolite,scales = "free_y")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  labs(title = "Longitudinal changes of the metabolites",
       x = "Timepoint",
       y = "Concentration",
       color = "SourceID")


```


## spectra pre-procssing

```{r}
spec_cal <- calibrateSpectra(
  ppm,
  spec,
  ref = c("alanine"),
  rOref = c(1.46, 1.50),
  cshift = 1.485,
  maxShift = 1.5
)

idx<-c(which(ppm<=0.5),which(ppm>=4.6 & ppm<=5.0),which(ppm>=9.5))

spec_cal<-spec_cal[,-idx]
ppm<-ppm[-idx]
rm(idx)
spec_bcor<-baselineCorrection(spec_cal)
spec_pqn<-normalize(spec_bcor)
colnames(spec_pqn) <- ppm
rownames(spec_pqn) <- spec_anno$title

```


## PCA analysis of NMR Data (untargeted approach)
```{r}
mod_nmr<-PCA(spec_pqn)

plotScores(model = mod_nmr,optns = list(color = factor(MetaData$SourceID),shape = MetaData$timepoint))

PlotLoadSpec(model = mod_nmr,roi = c(0.5,4.6),PC = 1)
PlotLoadSpec(model = mod_nmr,roi = c(5.0,9.5),PC = 1)

PlotLoadSpec(model = mod_nmr,roi = c(0.5,4.6),PC = 2)
PlotLoadSpec(model = mod_nmr,roi = c(5.0,9.5),PC = 2)

PlotLoadSpec(model = mod_nmr,roi = c(0.5,4.6),PC = 3)
PlotLoadSpec(model = mod_nmr,roi = c(5.0,9.5),PC = 3)
```


### APAP-Glucuronide

```{r}
matspec(spec_pqn,ppm,roi = c(7.12,7.17),interactive = T)
```



### APAP-Sulfate (and AS2)

```{r}
matspec(spec_pqn,ppm,roi = c(7.30,7.35),interactive = T) 
```


### APAP-Sulfate (and AS1)

```{r}
matspec(spec_pqn,ppm,roi = c(7.4,7.45),interactive = T) 
```

